\documentclass{article}


% This file is a solution template for:
\input{../preamble/lab_preamble.tex}


\title{Electronic Circuit Simulation and Layout Software}


\begin{document}
\maketitle

\section{Lab 6: Electronic Circuit Simulation and Layout Software}

\subsection*{General comments}

\begin{itemize}
\item 5Spice is only available for Windows.  It should be run as freeware.
\item EAGLE is available for Windows, Mac, Linux.  It should be run as freeware.  Installation in Windows requires that all previous versions are removed by hand first.
\end{itemize}

\subsection{5Spice}
\begin{itemize}
\item The freeware 5Spice license does not allow $x$ vs $y$ plots, nor does it allow copy \& paste or exporting of the table of results in a sweep.  Other solutions have to be found to achieve similar goals.
\item The diode model 1N4007 seems to have issues.
\item Simulations of high-frequency signals will need to use an explicit minimum step size at least 10 times smaller than the period of oscillation to avoid aliasing.
\item Measuring the ripple on the rise of a charging capacitor can be done by taking the average of the min/max differences for two consecutive minima around one maximum.  This should be compared to the steady-state voltage, though.
\end{itemize}

\subsection{EAGLE}
\begin{itemize}
\item Not terribly important to be detailed here, just get the students to do \emph{something}.
\item Discourage the students from using pre-packaged rectifiers.  Instead, they should use individual diodes.
\end{itemize}


\end{document}
